---
title: "Characterising Online Private Treaty Negotiations"
subtitle: "Field Evidence from Irish Residential Property"
author: "Damien Dupré and Healy Hynes"
format:
  revealjs:
    theme: [theme/metropolis.scss]
    toc-depth: 1
    chalkboard: false
    menu: true
    slide-number: true
    revealjs-plugins:
      - attribution
---

```{r setup, include=FALSE}
library(tidyverse)

# Load raw data
raw <- read_csv(
  "../../data/dcu_beagel_dataset_20251201.csv",
  show_col_types = FALSE
) |>
  rename(
    sale_id    = `Synthetic Sale ID`,
    bid_amount = `Bid Amount`,
    timestamp  = `Unix Timestamp`,
    bidder_id  = `Bidder ID`
  ) |>
  mutate(datetime = as.POSIXct(timestamp, origin = "1970-01-01", tz = "UTC"))

# Property-level: sort bids within each property by time
bids <- raw |>

  arrange(sale_id, timestamp)

# Compute consecutive bid increments and latencies (property level)
bids <- bids |>
  group_by(sale_id) |>
  mutate(
    bid_increment  = bid_amount - lag(bid_amount),
    time_diff_hrs  = as.numeric(difftime(datetime, lag(datetime), units = "hours"))
  ) |>
  ungroup()

# Property-level summaries
prop_summary <- bids |>
  group_by(sale_id) |>
  summarise(
    n_participants = n_distinct(bidder_id),
    n_bids         = n(),
    first_offer    = first(bid_amount),
    last_offer     = last(bid_amount),
    premium_pct    = (last_offer - first_offer) / first_offer * 100,
    .groups = "drop"
  )

# Filter to properties with 2+ participants
prop_summary <- prop_summary |>
  filter(n_participants >= 2)

# Participant-level summaries
# Identify each bidder's first bid position within a property
bidder_summary <- bids |>
  filter(sale_id %in% prop_summary$sale_id) |>
  group_by(sale_id) |>
  mutate(
    bid_order       = row_number(),
    total_bids_prop = n()
  ) |>
  ungroup() |>
  group_by(sale_id, bidder_id) |>
  summarise(
    avg_bid_increment = mean(bid_increment, na.rm = TRUE),
    avg_reaction_hrs  = mean(time_diff_hrs, na.rm = TRUE),
    n_bids            = n(),
    first_bid_order   = min(bid_order),
    total_bids_prop   = first(total_bids_prop),
    .groups = "drop"
  ) |>
  mutate(
    entry_stage  = first_bid_order / total_bids_prop,
    first_bidder = as.integer(first_bid_order == 1)
  )

# Common ggplot theme for slides
theme_presentation <- theme_minimal(base_size = 16) +
  theme(
    plot.title    = element_text(face = "bold", size = 18, hjust = 0.5),
    plot.subtitle = element_text(size = 14, hjust = 0.5, colour = "grey40"),
    panel.grid.minor = element_blank()
  )
```

## A Real Bidding War {background-color="#0b141a"}

<iframe data-src="whatsapp-bidding.html" width="150%" height="700" style="border:none;"></iframe>

## the Irish market is changing ...

- Irish property market shifting from **opaque, agent-mediated** negotiations to **transparent digital platforms**
- Online platforms allow buyers to observe competing bids in real time
- Perceived as **fairer** by a majority of the public (Robertson et al., 2026)
- But little is known on the impact of this new technology on the actual bidding behaviour.

::: footer
- [Robertson, D., Shier, A. J., Timmons, S., & Lunn, P. (2026). Buying and selling property in Ireland: Behavioural evidence on bidding systems, stressors and market comprehension. *ESRI Research Series No 226*](https://www.esri.ie/publications/buying-and-selling-property-in-ireland-behavioural-evidence-on-bidding-systems)
:::

## Data

- **Source:** Anonymised transactional logs from an online offer management platform in Ireland (2018–2024)
- **Each record**: timestamp, property-scoped buyer ID, offer amount (€), property ID
- **After filtering** (negotiations with 2+ participants):

```{r data-summary, echo=FALSE}
tibble(
  Metric = c("Properties", "Unique buyer", "Top-offer outcomes"),
  Value  = c(
    format(nrow(prop_summary), big.mark = ","),
    format(nrow(bidder_summary), big.mark = ","),
    paste0(round(mean(prop_summary$n_participants == 2) * 100), "%")
  )
) |>
  knitr::kable()
```

## Descriptive Analysis

**Property-level measures:**

- Number of active participants
- Total number of offers
- Duration of negotiation window (first to last offer)
- Trajectory of offer escalation

**Participant-level measures:**

- Average offer increment (€ above preceding offer)
- Average response latency (hours between offers)
- Number of offers placed

# Results

## Summary Statistics

```{r summary-stats, echo=FALSE}
clean_bidder <- bidder_summary |>
  filter(
    !is.na(avg_bid_increment) & is.finite(avg_bid_increment),
    !is.na(avg_reaction_hrs) & is.finite(avg_reaction_hrs)
  )

fmt <- function(x, digits) {
  sprintf(paste0("%.", digits, "f"), x)
}

entry_pct <- fmt(mean(clean_bidder$entry_stage, na.rm = TRUE) * 100, 0)
first_pct <- fmt(mean(clean_bidder$first_bidder, na.rm = TRUE) * 100, 0)

summary_tbl <- tibble(
  Variable = c(
    "Avg Bid Increment (\u20ac)",
    "Avg Reaction Speed (hrs)",
    "Total Bids Placed"
  ),
  Mean = c(
    fmt(mean(clean_bidder$avg_bid_increment, na.rm = TRUE), 0),
    fmt(mean(clean_bidder$avg_reaction_hrs, na.rm = TRUE), 2),
    fmt(mean(clean_bidder$n_bids, na.rm = TRUE), 2)
  ),
  SD = c(
    fmt(sd(clean_bidder$avg_bid_increment, na.rm = TRUE), 0),
    fmt(sd(clean_bidder$avg_reaction_hrs, na.rm = TRUE), 2),
    fmt(sd(clean_bidder$n_bids, na.rm = TRUE), 2)
  ),
  Min = c(
    fmt(min(clean_bidder$avg_bid_increment, na.rm = TRUE), 0),
    fmt(min(clean_bidder$avg_reaction_hrs, na.rm = TRUE), 2),
    fmt(min(clean_bidder$n_bids, na.rm = TRUE), 0)
  ),
  Max = c(
    fmt(max(clean_bidder$avg_bid_increment, na.rm = TRUE), 0),
    fmt(max(clean_bidder$avg_reaction_hrs, na.rm = TRUE), 2),
    fmt(max(clean_bidder$n_bids, na.rm = TRUE), 0)
  )
)

knitr::kable(summary_tbl, format = "markdown", align = "lrrrr")
```

On average, bidders entered at **`r entry_pct`%** of the way through the bidding sequence. **`r first_pct`%** of bidders placed the first bid on a property.

## Correlations

:::: {layout="[2,1]"}
::: {#first-column}
```{r plot-correlations, echo=FALSE, fig.width=7, fig.height=5.5}
cor_data <- clean_bidder |>
  select(
    `Avg Bid\nIncrement`  = avg_bid_increment,
    `Avg Reaction\nSpeed` = avg_reaction_hrs,
    `Entry\nStage`        = entry_stage,
    `First\nBidder`       = first_bidder,
    `Total\nBids`         = n_bids
  )

cor_mat <- cor(cor_data, use = "pairwise.complete.obs")

# Keep only the lower triangle (set upper triangle to NA)
cor_mat[upper.tri(cor_mat)] <- NA

cor_long <- as.data.frame(as.table(cor_mat)) |>
  rename(Var1 = Var1, Var2 = Var2, value = Freq) |>
  filter(!is.na(value)) |>
  mutate(
    label = ifelse(
      round(abs(value), 2) == 1,
      "1.00",
      sub("^0", "", sprintf("%.2f", value))
    )
  )

ggplot(cor_long, aes(x = Var2, y = fct_rev(Var1), fill = value)) +
  geom_tile(colour = "white", linewidth = 1) +
  geom_text(aes(label = label), size = 5, fontface = "bold") +
  scale_fill_gradient2(
    low = "#e74c3c", mid = "white", high = "#1a5276",
    midpoint = 0, limits = c(-1, 1),
    name = "r"
  ) +
  labs(
    title = "Correlation Matrix of\nNegotiation Variables",
    x = NULL, y = NULL
  ) +
  theme_presentation +
  theme(
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.position = "right"
  ) +
  coord_fixed()
```
:::

::: {#second-column}
- Later entrants bid higher and respond more slowly. 

- First bidders place more offers but with smaller increments.
:::
::::

## Competition

:::: {layout="[2,1]"}
::: {#first-column}
```{r plot-competition, echo=FALSE, fig.width=7, fig.height=5}
ggplot(prop_summary, aes(x = n_participants)) +
  geom_histogram(
    fill = "#2980b9", alpha = 0.85, colour = "white",
    binwidth = 1, boundary = 0.5
  ) +
  geom_vline(
    aes(xintercept = mean(n_participants)),
    colour = "#e74c3c", linewidth = 1, linetype = "dashed"
  ) +
  annotate(
    "text",
    x = mean(prop_summary$n_participants) * 1.15,
    y = Inf, vjust = 2, hjust = 0,
    label = paste0("Mean \u2248 ", round(mean(prop_summary$n_participants), 1), " participants"),
    colour = "#e74c3c", fontface = "bold", size = 5
  ) +
  scale_x_continuous(breaks = seq(2, max(prop_summary$n_participants), 1)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Distribution of Active Participants\nper Property",
    x = "Number of Active Participants",
    y = "Number of Properties"
  ) +
  theme_presentation
```
:::

::: {#second-column}
- Most properties attract only 2 or 3 active bidders. 

- However, a long tail of highly contested negotiations exists.
:::
::::

## Offer Increments

:::: {layout="[2,1]"}
::: {#first-column}
```{r plot-increments, echo=FALSE, fig.width=7, fig.height=5}
increments <- bidder_summary |>
  filter(!is.na(avg_bid_increment) & is.finite(avg_bid_increment) & avg_bid_increment > 0)

ggplot(increments, aes(x = avg_bid_increment)) +
  geom_histogram(
    fill = "#27ae60", alpha = 0.85, colour = "white", bins = 30
  ) +
  geom_vline(
    aes(xintercept = mean(avg_bid_increment)),
    colour = "#e74c3c", linewidth = 1, linetype = "dashed"
  ) +
  annotate(
    "text",
    x = mean(increments$avg_bid_increment) * 1.1,
    y = Inf, vjust = 2, hjust = 0,
    label = paste0(
      "Mean = \u20ac", format(round(mean(increments$avg_bid_increment)), big.mark = ","),
      "\nSD = \u20ac", format(round(sd(increments$avg_bid_increment)), big.mark = ",")
    ),
    colour = "#e74c3c", fontface = "bold", size = 4.5
  ) +
  scale_x_log10(
    labels = scales::label_currency(prefix = "\u20ac", big.mark = ",")
  ) +
  labs(
    title = "Distribution of Average Offer\nIncrements",
    x = "Average Offer Increment (\u20ac, log scale)",
    y = "Frequency"
  ) +
  theme_presentation
```
:::

::: {#second-column}
- Bid increments are right-skewed as most are modest, but some buyers make very large jumps to signal commitment. 

- Heavy use of €1k, €2.5k, and €5k jumps.
:::
::::

## Response Latency

:::: {layout="[2,1]"}
::: {#first-column}
```{r plot-latency, echo=FALSE, fig.width=7, fig.height=5}
latency <- bidder_summary |>
  filter(!is.na(avg_reaction_hrs) & is.finite(avg_reaction_hrs) & avg_reaction_hrs > 0)

ggplot(latency, aes(x = avg_reaction_hrs)) +
  geom_histogram(
    fill = "#8e44ad", alpha = 0.85, colour = "white", bins = 40
  ) +
  geom_vline(
    aes(xintercept = mean(avg_reaction_hrs)),
    colour = "#e74c3c", linewidth = 1, linetype = "dashed"
  ) +
  annotate(
    "text",
    x = mean(latency$avg_reaction_hrs) * 1.3,
    y = Inf, vjust = 2, hjust = 0,
    label = paste0(
      "Mean \u2248 ", round(mean(latency$avg_reaction_hrs), 1), " hours",
      "\nSD = ", round(sd(latency$avg_reaction_hrs), 1), " hours"
    ),
    colour = "#e74c3c", fontface = "bold", size = 4.5
  ) +
  scale_x_continuous(
    trans = "log1p",
    breaks = c(1, 6, 12, 24, 48, 168, 720),
    labels = c("1h", "6h", "12h", "1d", "2d", "1w", "1m")
  ) +
  labs(
    title = "Distribution of Average\nResponse Latency",
    x = "Average Response Latency (log scale)",
    y = "Frequency"
  ) +
  theme_presentation
```
:::

::: {#second-column}
Most responses come within hours, suggesting real-time competitive pressure, though some negotiations stretch over weeks.
:::
::::

## Competition and Price Premium

:::: {layout="[2,1]"}
::: {#first-column}
```{r plot-premium, echo=FALSE, fig.width=7, fig.height=5}
premium_ind <- prop_summary |>
  mutate(
    n_grp = case_when(
      n_participants == 2 ~ "2",
      n_participants == 3 ~ "3",
      n_participants == 4 ~ "4",
      n_participants >= 5 ~ "5+"
    ),
    n_grp = factor(n_grp, levels = c("2", "3", "4", "5+")),
    n_num  = as.numeric(n_grp)
  )

# Mean and upper whisker per group (for label placement)
premium_means <- premium_ind |>
  group_by(n_grp, n_num) |>
  summarise(
    mean_premium = mean(premium_pct, na.rm = TRUE),
    q3           = quantile(premium_pct, 0.75, na.rm = TRUE),
    iqr          = IQR(premium_pct, na.rm = TRUE),
    upper_whisker = min(
      max(premium_pct, na.rm = TRUE),
      q3 + 1.5 * iqr
    ),
    .groups = "drop"
  )

ggplot(premium_ind, aes(x = n_num, y = premium_pct)) +
  # Half-violin (density) -- shifted right
  geom_violin(
    aes(group = n_grp),
    trim = TRUE, scale = "width", width = 0.6,
    fill = "#e67e22", alpha = 0.3, colour = NA,
    position = position_nudge(x = 0.15)
  ) +
  # Boxplot -- centred
  geom_boxplot(
    aes(group = n_grp),
    width = 0.15, outlier.shape = NA,
    fill = "#e67e22", alpha = 0.7, colour = "#2c3e50"
  ) +
  # Jittered points -- shifted left
  geom_jitter(
    aes(x = n_num - 0.2),
    width = 0.06, height = 0, size = 1.2,
    alpha = 0.35, colour = "#c0392b"
  ) +
  # Mean markers
  geom_point(
    data = premium_means,
    aes(x = n_num, y = mean_premium),
    shape = 18, size = 4, colour = "#2c3e50"
  ) +
  # Mean labels -- positioned above the upper whisker to avoid overlap
  geom_text(
    data = premium_means,
    aes(
      x = n_num, y = upper_whisker,
      label = paste0("M = ", round(mean_premium, 1), "%")
    ),
    vjust = -0.8, fontface = "bold", colour = "#2c3e50", size = 4.5
  ) +
  scale_x_continuous(
    breaks = 1:4,
    labels = levels(premium_ind$n_grp)
  ) +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Competition Intensity and\nPrice Premium",
    x = "Number of Active Participants",
    y = "Premium Above First Offer (%)"
  ) +
  theme_presentation +
  theme(plot.margin = margin(15, 10, 10, 10))
```
:::

::: {#second-column}
More participants drive higher final prices, consistent with competitive arousal and auction fever dynamics.
:::
::::

<!-- ## Ghost Bidders or Chancers -->

<!-- :::: {layout="[2,1]"} -->
<!-- ::: {#first-column} -->
<!-- ```{r plot-ghost, echo=FALSE, fig.width=7, fig.height=5} -->
<!-- # A "ghost bidder" places exactly 1 bid and never returns, -->
<!-- # EXCLUDING single-bid participants whose bid was the top offer -->
<!-- top_bidders <- bids |> -->
<!--   filter(sale_id %in% prop_summary$sale_id) |> -->
<!--   group_by(sale_id) |> -->
<!--   slice_max(bid_amount, n = 1, with_ties = FALSE) |> -->
<!--   ungroup() |> -->
<!--   select(sale_id, top_bidder = bidder_id) -->

<!-- ghost_grouped <- bidder_summary |> -->
<!--   left_join( -->
<!--     prop_summary |> select(sale_id, n_participants), -->
<!--     by = "sale_id" -->
<!--   ) |> -->
<!--   left_join(top_bidders, by = "sale_id") |> -->
<!--   mutate( -->
<!--     is_top_bidder = (bidder_id == top_bidder), -->
<!--     n_grp = case_when( -->
<!--       n_participants == 2 ~ "2", -->
<!--       n_participants == 3 ~ "3", -->
<!--       n_participants == 4 ~ "4", -->
<!--       n_participants >= 5 ~ "5+" -->
<!--     ), -->
<!--     n_grp = factor(n_grp, levels = c("2", "3", "4", "5+")), -->
<!--     bidder_type = case_when( -->
<!--       n_bids == 1 & is_top_bidder  ~ "Ghost + Top Offer", -->
<!--       n_bids == 1 & !is_top_bidder ~ "Ghost (not top)", -->
<!--       TRUE ~ "Active (2+ bids)" -->
<!--     ), -->
<!--     bidder_type = factor( -->
<!--       bidder_type, -->
<!--       levels = c("Ghost + Top Offer", "Ghost (not top)", "Active (2+ bids)") -->
<!--     ) -->
<!--   ) |> -->
<!--   count(n_grp, bidder_type, .drop = FALSE) |> -->
<!--   group_by(n_grp) |> -->
<!--   mutate(pct = n / sum(n) * 100) |> -->
<!--   ungroup() -->

<!-- ggplot(ghost_grouped, aes(x = n_grp, y = n, fill = bidder_type)) + -->
<!--   geom_col(position = position_dodge(width = 0.8), width = 0.7, alpha = 0.85) + -->
<!--   geom_text( -->
<!--     aes(label = paste0(round(pct), "%")), -->
<!--     position = position_dodge(width = 0.8), -->
<!--     vjust = -0.5, fontface = "bold", size = 4 -->
<!--   ) + -->
<!--   scale_fill_manual( -->
<!--     values = c( -->
<!--       "Ghost + Top Offer" = "#e67e22", -->
<!--       "Ghost (not top)"   = "#7f8c8d", -->
<!--       "Active (2+ bids)"  = "#2980b9" -->
<!--     ), -->
<!--     name = NULL -->
<!--   ) + -->
<!--   scale_y_continuous(expand = expansion(mult = c(0, 0.15))) + -->
<!--   labs( -->
<!--     title = "Ghost vs Active Bidders by\nCompetition Intensity", -->
<!--     subtitle = "Ghost = placed exactly one bid", -->
<!--     x = "Number of Active Participants", -->
<!--     y = "Number of Bidder-Property Pairs" -->
<!--   ) + -->
<!--   theme_presentation + -->
<!--   theme(legend.position = "top") -->
<!-- ``` -->
<!-- ::: -->

<!-- ::: {#second-column} -->
<!-- A substantial share of participants bid once and vanish, inflating apparent competition without sustained engagement. -->
<!-- ::: -->
<!-- :::: -->

## Discussion & Implications

- Real-world online negotiations involve **meaningful competitive dynamics**, not just passive price discovery
- The association between participation intensity and premiums is consistent with **competitive arousal** (McGee, 2013; Hammervold et al., 2025)
- **Future work:** examination of offer-level behaviours associated with different negotiation outcomes

::: footer
- [McGee, P. (2013). Bidding in private-value auctions with uncertain values. *Games and Economic Behavior*, 82, 312–326.](https://www.sciencedirect.com/science/article/abs/pii/S0899825613001188)
- [Hammervold, R., Olaussen, J. O., Oust, A., Sommervoll, D. E., & Sønstebø, O. J. (2025). Some like it hot: The effect of auction temperature on urban real estate prices. Real Estate Economics, 53, 1284–1308.](https://doi.org/10.1111/1540-6229.70004)

:::

<!-- ## References -->

<!-- ::: {style="font-size: 0.55em;"} -->

<!-- - Hammervold, R. et al. (2025). Some like it hot: The effect of auction temperature on urban real estate prices. *Real Estate Economics*, 53(6), 1284–1308. -->
<!-- - Kahneman, D. & Tversky, A. (1979). Prospect theory. *Econometrica*, 47(2), 263–291. -->
<!-- - Lunn, P. et al. (2025). Buying and selling houses in Ireland. *ESRI Working Paper*, No. 811. -->
<!-- - McGee, P. (2013). Bidding in private-value auctions with uncertain values. *Games and Economic Behavior*, 82, 312–326. -->
<!-- - Robertson, D. et al. (2026). Buying and selling property in Ireland. *ESRI Research Series*, No. 226. -->
<!-- - Sonstebo, O.J. et al. (2021). The heat of the moment. *Scandinavian Journal of Economics*, 123(4), 1188–1215. -->

<!-- ::: -->

## {background-color="#0b141a"}

<iframe data-src="thankyou-phone.html" width="100%" height="700" style="border:none;"></iframe>
